---
title: Table of Model Results using kable and kableExtra
author: Daniel Chen
date: '2019-09-13'
slug: table-of-model-results-using-kable-and-kableextra
categories: []
tags:
  - R
  - rmarkdown
subtitle: ''
summary: ''
authors: []
lastmod: '2019-09-13T10:30:59-04:00'
featured: no
projects: []
---

I'm at the [R/Medicine][rmedicine] conference (no it's not a Reddit thing) and
got to help Alison Hill with her [R Markdown for Medicine][rmd4medicine] workshop.
One of the questions I got to tinker with was making tables used to report model results.

<!-- more -->

One technique I learned while doing my MPH was to add variables to your model in blocks.
It reduces the number of tests you need to perform, and is more meaningful than saying "I ran stepwise".
So, as a researcher, you will add variables in meaningful blocks and then see if that block of variables
is "significant" by looking at the F-statistic.

Here is one example of at least putting your model results into a table.

## Creating the models

```{r}
library(reshape2) # using this just for the tips dataset
library(broom)
library(purrr)
library(knitr)
library(kableExtra)

head(tips)

# create a bumch of models
m1 <- lm(tip ~ total_bill, data = tips)
m2 <- lm(tip ~ total_bill + sex, data = tips)
m3 <- lm(tip ~ total_bill + sex + smoker, data = tips)
m4 <- lm(tip ~ total_bill + smoker + size, data = tips)
```

Now that we have the models, we can use the `broom` package to get a nice consistent dataframe of the coefficients

```{r}
# get a dataframe of the model estimates
models <- list(m1, m2, m3, m4)
broom_outputs <- purrr::map(models, broom::tidy)
broom_outputs
```

What we have now is a list of dataframes, one for each model we fit.

## Just the beta estimates

If we just want the coefficients, we only need the `terms` and `estimates` columns and do a `join`.
We can use the `purrr::reduce` function to do this

```{r}
# join them all together
model_outputs <- purrr::reduce(broom_outputs, dplyr::full_join, by = 'term')
model_outputs
```

This will join all the dataframes together using the `term` column as the key.
Variables that do not exist in the model will be filled with an `NA`.
Problem now, is that we have a bunch of columns we don't need.
So the next step is to pull out the columns we want, and rename them accordingly.

```{r}
model_table <- model_outputs %>%
  dplyr::select(term, tidyselect::starts_with('estimate'))
```

```{r}
names(model_table) <- c('Variable', 'Model 1', "Model 2", "Model 3", "Model 4")
model_table
```


## What if I need more variables?

Just reporting the beta coefficients is not how you want to report your findings,
typically you'd want to also display some measure of uncertainty along with the results.
Here I'm just going to add the `std.error` column, but you can calculate the 95% confidence intervals using the `std.error`.

Here's the reference from the `kableExtra` library I'm using to build the coefficient table:

```{r}
# this is the exmaple form the kable docs I used for reference
# https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html

collapse_rows_dt <- data.frame(C1 = c(rep("a", 3), rep("b", 2)),
                               C2 = c(rep("c", 2), rep("d", 1), rep("c", 1), rep("d", 1)),
                               C3 = 1:5,
                               C4 = sample(c(0,1), 5, replace = TRUE))
collapse_rows_dt
```

```{r}
kable(collapse_rows_dt, align = "c") %>%
  kable_styling(full_width = F) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:2, valign = "top")
```

This example was important to my solution, becuase I would need to format my coefficient table the same way as `collapse_rows_dt`.
This means that:

- `C1` would be the variables in my model
- `C2` would need to alternate between the `estimate` and the `std.error`
- The other columns would be the actual values for each model.

Since the model outputs the `estimate` and `std.error` as columns,
I know some kind of reshaping/tidying step will need to be involved (`melt`/`gather` / `pivot_longer`).

We still start from the same data steps as the previous example

```{r}
broom_outputs
```

The trick here is to get the data in the correct format that can be used by `kableExtra`

```{r}
broom_stats_long <- purrr::map(broom_outputs, tidyr::gather, 'key', 'value', estimate, std.error)
broom_stats_long
```

Now that we have the data formatted the way we need, we now `select` only the columns we need.

```{r}
# filter only the columns we need
broom_stats_long_filter <- purrr::map(broom_stats_long, dplyr::select, term, key, value)
broom_stats_long_filter
```

Next, you need to create some kind of ID value for each model you're creating,
that way when you're combining everything into a single dataframe,
the estimates are traveling together with the standard error.
Here I'm labeling each model as a number starting with `1`.

```{r}
tidy_tables <- purrr::map2_df(broom_stats_long_filter, 1:length(broom_stats_long_filter),
                              function(x, y) dplyr::mutate(x, mod = y))
tidy_tables
```

Almost there!

Lastly, we need to sort the columns by using `arrange`.
Here is where the ID column, `mod` get's used to make sure model values are traveling together.

```{r}
df <- tidy_tables %>%
  dplyr::arrange(term, mod, key, value) %>%
  tidyr::spread(mod, value, drop = FALSE)
df
```

```{r}
kable(df, align='c') %>%
  kable_styling(full_width = F) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1, valign = "top")
```


## kableExtra

There kableExtra is great for tables like this. Here are links to more information

- https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html
- http://haozhu233.github.io/kableExtra/

[rmedicine]: https://r-medicine.com/
[rmd4medicine]: https://rmd4medicine.netlify.com/
