---
title: 'Plotting Ebola: Collaboration in Open Science'
author: "Daniel Chen"
date: "11/21/2014"
output:
  ioslides_presentation: default
  beamer_presentation: default
  slidy_presentation: default
---

# Open Science & Open Source

## Data Sharing and Management Snafu in 3 Short Acts

http://www.youtube.com/watch?v=N2zK3sAtr-4

## Recap

- I am a researcher who wants to ask a research question using someone else's data
- The data software is no longer in development/maintained
- There is no code book
- The variables do not make sense

## Tools
- IPython Notebook (Python)
- Knitr/sweave documents (R)
    - Fun fact: this presentation is written in a plain text file in R/RStudio
- Any means that can interweave text, code, and ouput
    - org files

## Open Science
- Enables true reproducible research
    - Makes data analysis easier to record, understand and reproduce
    - Detailed records of their work
    - Share
- Enables collaboration
    - Devise teaching modules and collaborate with others

## Journals
- PLOS ONE
- Nature
    - [Share](http://www.nature.com/ismej/journal/v7/n3/full/ismej2012123a.html) the code! 
    - [Collaborative cloud-enabled tools allow rapid, reproducible biological insights](http://www.nature.com/ismej/journal/v7/n3/full/ismej2012123a.html)

## Willingness to Share Research Data Is Related to the Strength of the Evidence and the Quality of Reporting of Statistical Results
J.M. Wicherts, M. Bakker, and D. Molenaar: "Willingness to Share Research Data Is Related to the Strength of the Evidence and the Quality of Reporting of Statistical Results". PLoS ONE, 6(11): e26828, 2011, [`doi:10.1371/journal.pone.0026828`](http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0026828).

## Open Source
- R and Python
- Large community and user base
- Good developers
- Get your research out there faster
- People can use it for free

## What I've learned from Greg Wilson and Software Carpentry
- Most scientists think of programming as a tax they have to pay in order to do science
    - get that piece of lab equiptment to work
    - get data
    - get paper
    - get sleep

Wilson G. Software Carpentry: lessons learned [v1; ref status: indexed, http://f1000r.es/2x7] F1000Research 2014, 3:62 (doi: 10.12688/f1000research.3-62.v1) - See more at: http://f1000research.com/articles/3-62/v1#sthash.LbJrvdrL.dpuf

# Ebola
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(RCurl)
library(reshape2)
library(magrittr)
library(dplyr)
library(RColorBrewer)
library(ggplot2)
```

```{r, cache=TRUE, echo=FALSE}
url <- "https://raw.githubusercontent.com/cmrivers/ebola/master/country_timeseries.csv"

data <- getURL(url, ssl.verifypeer = FALSE)
df <- read.csv(textConnection(data))
```

## The data
Case counts and deaths is from the World Health Organization and WHO situation reports
```{r}
latest <- head(df, n = 2) %>% t() %>% as.data.frame()
pattern = '^Date|^Day|^Cases'
rows = grep(pattern = pattern, x = row.names(latest))
latest[rows, ]
```

## Data munging
```{r}
# drop the date
df1_noDate <- df[, !names(df) %in% c("Date")]
```

```{r}
# Shift dates so they are all relative to eachother
# Build a series from 0...latest day in data set
day <- c(0:max(df1_noDate$Day))
# We'll add updates on each day we have data for each country here
df3_merge <- data.frame(day)
```

## Data munging
```{r}
# For each country:
for(country in 2:ncol(df1_noDate)){
  df_temp <- df1_noDate[, c(1, country)]
  # Data set is snapshots at day of reporting, 
  # with NAs representing "no change"/"no new data"
  # so ignore those with NAs.
  df_temp <- na.omit(df_temp)

  #Rescale all series so day 0 == first reported case/death
  df_temp$day.adj <- df_temp$Day - min(df_temp$Day)

  df3_merge <- merge(x = df3_merge, 
                     y = df_temp[, names(df_temp) != "Day"],
                     by.x = "day", by.y = "day.adj", all.x = TRUE)
}
```

## Data munging
```{r}
# assign day to row.names
row.names(df3_merge) <- df3_merge$day
# drop 'day'
df3_merge <- df3_merge[, names(df3_merge) != "day"]
head(df3_merge, n = 1)
```

## Data munging
```{r}
# transpose
df4 <- as.data.frame(t(as.matrix(df3_merge)))
head(df4[, c(1:10)])
```

```{r}
vars <- colsplit(row.names(df4), "_", c("type", "place"))
df4 <- cbind(vars, df4)
row.names(df4) <- NULL
```

## Data munging
```{r]}
df5_melt <- melt(df4)
names(df5_melt) <- c("type", "place", "day", "count")
df5_melt$type[df5_melt$type == "Case"] <- "Cases"
head(df5_melt)
```

## Variables for plotting
```{r}
all <- unique(df5_melt$place)
c_colors <- brewer.pal(length(all), 'Set1')
names(c_colors) <- all

theme_set(theme_minimal())

# drop NA observations for plotting
df_plot <- df5_melt[!is.na(df5_melt$count), ]
selection <- all

df_plot <- df_plot %>% 
    filter(place %in% selection) %>%
    mutate(count = as.numeric(count), day=as.numeric(day))

data_plot <- df_plot
```

## Data for plotting
```{r}
head(data_plot)
```


## Initial Plot
```{r}
g <- ggplot(data = data_plot,
            aes(x = day, y = count,
                group = place, color = place)) +
    geom_point() + geom_line()+
    facet_grid(~ type) +
    scale_x_continuous(name="Days rom index") +
    scale_y_continuous(name="Counts") +
    scale_colour_manual(name="Country", values=c_colors) +
    ggtitle("Number of observations for days after first report")
```

## Plot
```{r}
g
```



## Deployed Application

https://chendaniely.shinyapps.io/shinyCountryTimeseries

## The repositories

This is all possible beacause of [Caitlin](http://www.caitlinrivers.com/) [Rivers](https://twitter.com/cmyeaton) and all the [contributors](https://github.com/cmrivers/ebola/graphs/contributors) (25+) to the [repository](https://github.com/cmrivers/ebola)

    - Computational epidemiology
    - Network Dynamics and Simulation Science Laboratory (NDSSL) 
        - Virginia Bioinformatics Institute, Virginia Tech

My [fork](https://github.com/chendaniely/ebola) is the staging area for the shiny application

## Shiny contributors
- Timothée Poisot [tpoisot](https://github.com/tpoisot)
    - Québec Centre for Biodiversity Science, Rimouski, Québec, Canada
- Mathew Hall [mathew-hall](https://github.com/mathew-hall/)
    - Sheffield, England
- Ryan Quan [rcquan](https://github.com/rcquan/)
    - Columbia University Mailman School of Public Health, New York NY

## Why
Our planet needs our help, and we need (good) science to fix it.

# Thanks
Twitter: [chendaniely](https://twitter.com/chendaniely)
Github: [chendaniely](https://github.com/chendaniely)

Open Source Report Card: https://osrc.dfm.io/
